---
title: "pathogen_study"
author: "Becky S"
date: "2026-05-21"
output: word_document
---

## Import Data & create dataframe

1.  Import from pathogen data, but for now exclude c.jejuni, c.lari and 16scampy
2.  Add pathogen indicators together to create new variable: pathogens_add
3.  Remove individual pathogen columns
4.  Create dataframe, apply log transform on all data, and if indicators contain "-88", convert to NA to exclude from regression analysis

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)

#import data:
readRenviron(".Renviron")
data_path <- Sys.getenv("DATA_PATH")

data<-read_csv(file.path(data_path, "pathogen_data.csv")) %>% 
  select(-c("c.coli","c.jejuni","c.lari", "16scampy"))

df_1 <- data %>% 
  dplyr::mutate(norog1 = ifelse(norog1 == -88, 0, norog1)) %>% 
  dplyr::mutate(norog2 = ifelse(norog2 == -88, 0, norog2)) %>% 
  dplyr::mutate(adv = ifelse(adv == -88, 0, adv)) %>% 
  dplyr::mutate(pathogens_add = norog1 + norog2 + adv) %>% 
  select(-norog1, -norog2, -adv) %>% 
  dplyr::mutate(pathogens_add = ifelse(pathogens_add == 0, NA, pathogens_add)) %>% 
  select(-SampleDate, -StationCode)

df1 <- as.matrix(df_1)

df_1_log <- log(df1)
df_1_log[!is.finite(df_1_log)] <- NA
```

#non-parametric statistics:

```{r}
library("Hmisc") #for rcorr
spearman_rcorr <- rcorr(df_1_log, type = "spearman")

P_values <- as.data.frame(spearman_rcorr$P) 
R_values <- as.data.frame(spearman_rcorr$r) 
n_values <- as.data.frame(spearman_rcorr$n) 

print(P_values)
print(R_values)
print(n_values)

```

# pathogens vs. HF183

Explore the relationship between pathogens and HF183 as an indicator Note that HF183 is an indicator of human fecal bacteria, but it is not an indicator of a pathogen. So you may find HF183 and no-pathogens.

Uncertainties: HF183 is far less stable than pathogens in the environment, so could you could detect a low concentration of HF183 if the material has been in the environment for a long time, despite potentially having high amounts of pathogens

## basic biplot and distribution:

```{r, echo=FALSE}
#look at data  (parametric evaluation as first step)
HF183 <- as.data.frame(df_1_log) %>% 
  select(HF183, pathogens_add)

#simple linear regression (this model assumes a parametric relationship, but histogram shows non-normal distribution)
print(hist(HF183$HF183))

#confirm non-normality using Shapiro-Wild test
type_check <- shapiro.test(HF183$HF183)
print(type_check)
#if p-value is < 0.05, sufficient evidence that sample data does not come from a normal distribution

#visually inspect basic biplot and linear model (parametric)
library(ggplot2)
lm_HF183 <- lm(HF183$HF183 ~ HF183$pathogens_add)
summary(lm_HF183)
p1 <- ggplot(HF183, aes(x = pathogens_add, y = HF183))+
  geom_point(color= "blue", size= 2)+
  geom_smooth(method = "lm", se= TRUE, color = "red", linetype = "solid")+
  labs(x = "log(pathogen)",
       y = "log(HF183)")+
  theme_minimal()
print(p1)

#stats for lm
lm_HF183 <- lm(HF183$HF183 ~ HF183$pathogens_add)
summary(lm_HF183)

#output: lm p-value 0.101

```

## Spearman relationship

```{r}
HF183 <- as.data.frame(df_1_log) %>% 
  select(HF183, pathogens_add)

spearman_corr <- cor.test(HF183$HF183, HF183$pathogens_add, method = "spearman")
print(spearman_corr)

#data contains ties (multiple observations with the same value, preventing assignment of unique ranks for spearman- resulting p-value approximate)

#p-value: 0.098


```

# pathogens vs. crassphage

```{r}
#look at data  (parametric evaluation as first step)
crass <- as.data.frame(df_1_log) %>% 
  select(crassphage, pathogens_add)

#simple linear regression (this model assumes a parametric relationship, but histogram shows non-normal distribution)
hist(crass$crassphage)

#confirm non-normality using Shapiro-Wild test
type_check <- shapiro.test(crass$crassphage)
print(type_check)
#if p-value is < 0.05, sufficient evidence that sample data does not come from a normal distribution

#visually inspect basic biplot
library(ggplot2)

p2 <- ggplot(crass, aes(x = pathogens_add, y = crassphage))+
  geom_point(color= "blue", size= 2)+
  geom_smooth(method = "lm", se= TRUE, color = "red", linetype = "solid")+
  labs(x = "log(pathogen)",
       y = "log(HF183)")+
  theme_minimal()
print(p2)
```
